---
title: "LSI and clustering of tile matrix"
output: html_notebook
---

# Combine projects for all samples

```{r}
proj1a = loadArchRProject("../m154207/ArchR_bap2/ArchROutput/")
proj1b = loadArchRProject("../m154211/ArchR_bap2/ArchROutput/")
proj1c = loadArchRProject("../m167108/ArchR_bap2/ArchROutput/")
proj1d = loadArchRProject("../m168101/ArchR_bap2/ArchROutput/")
proj1e = loadArchRProject("../m167203/ArchR_bap2/ArchROutput/")

proj1 = ArchRProject(
  ArrowFiles = c(
    "../m154207/ArchR_bap2/ArchROutput/ArrowFiles/m154207.arrow",
    "../m154211/ArchR_bap2/ArchROutput/ArrowFiles/m154211.arrow",
    "../m167108/ArchR_bap2/ArchROutput/ArrowFiles/m167108.arrow",
    "../m168101/ArchR_bap2/ArchROutput/ArrowFiles/m168101.arrow",
    "../m167203/ArchR_bap2/ArchROutput/ArrowFiles/m167203.arrow"),
  copyArrows = FALSE,
  geneAnnotation = geneAnnotation,
  genomeAnnotation = genomeAnnotation,
  showLogo = FALSE
)

proj1 = proj1[
  which(proj1$cellNames %in%
          c(proj1a$cellNames,
            proj1b$cellNames,
            proj1c$cellNames,
            proj1d$cellNames,
            proj1e$cellNames)), ]
table(proj1$Sample)
getAvailableMatrices(proj1)

rm(proj1a, proj1b, proj1c, proj1d, proj1e)
```

Load ploidy computed by scPloidy

```{r}
ploidy =
  do.call(
    rbind,
    lapply(
      unique(proj1$Sample),
      function (s) {
        ploidy =
          read.table(
            paste0("../",
                   s,
                   "/ArchR_bap2/ploidy.txt"),
            header = TRUE,
            sep = "\t")
        ploidy$barcode =
          paste0(s, "#possorted_bam_", ploidy$barcode)
        return(ploidy)
      }))
all(proj1$cellNames %in% ploidy$barcode)
proj1$ploidy =
  ploidy$ploidy[match(proj1$cellNames, ploidy$barcode)]
rm(ploidy)
```

# Loading and QC of tile matrix (1st round)

Load tile matrix

```{r}
mat =
  getMatrixFromProject(
    proj1,
    useMatrix = "TileMatrix",
    binarize = TRUE)
mat$ploidy =
  proj1$ploidy[match(colnames(mat), proj1$cellNames)]
```

QC

```{r}
# colSm will not be updated. To be used later in QC plot.
colSm <- Matrix::colSums(assay(mat))

rowSm <- Matrix::rowSums(assay(mat))

# Clean up zero/singleton rows
# Actually redundant, because later filtered by adjvar
idx <- which(rowSm > 1)
mat <- mat[idx, ]
rowSm <- rowSm[idx]

# Exclude a small number of ~100% inserted windows
# Actually redundant, because later filtered by adjvar
idx <- which(rowSm/ncol(mat) < 0.85)
mat <- mat[idx, ]
rowSm <- rowSm[idx]
```

# LSI (1st round)

Smaller submatrix for faster computation

```{r}
adjvarmin = 0.4

# variance after IDF weighting
adjvar = log2(ncol(mat)/rowSm)^2 *
  (rowSm/ncol(mat)) *
  (1 - rowSm/ncol(mat))

qplot(
  x = rowSm/ncol(mat),
  y = adjvar,
  geom = "bin2d")
quantile(adjvar, seq(0, 1, 0.1))

idx = which(adjvar > adjvarmin)
mat2 = mat[idx, ]
rowSm2 = rowSm[idx]
```

Library size correct to unit vector

```{r}
assay(mat2) = assay(mat2) %*%
  Diagonal(x = 1/sqrt(colSums(assay(mat2)^2)))
```

Sample batch correct mat2

```{r}
target = rowMeans(assay(mat2))

sampleLabels = unique(as.character(mat$Sample))
for (s in sampleLabels) {
  print(s)
  x = assay(mat2) %*% Diagonal(x = as.numeric(mat2$Sample == s))
  y = assay(mat2) %*% Diagonal(x = as.numeric(mat2$Sample != s))
  rowMeansx = rowSums(x) / sum(mat2$Sample == s)
  correct = target / rowMeansx
  correct[rowMeansx == 0] = 1
  correctcap =
    quantile(correct, 0.75, na.rm = TRUE)^2 /
    quantile(correct, 0.5, na.rm = TRUE)
  correct = pmin(correct, correctcap)
  x = Diagonal(x = correct) %*% x

  assay(mat2) = x + y
  rm(x, y, rowMeansx, correct, correctcap)
}
```

IDF (inverse document frequency)
rowSm2 is calculated from original binary matrix

```{r}
assay(mat2) = Diagonal(x = log2(ncol(mat2) / rowSm2)) %*%
  assay(mat2)
```

Row centralization is mandatory!
[Skillicorn, Understanding complex datasets. p.51]

```{r}
gc()
set.seed(1)

# irlba only provides column centralization, not for rows
svd <- irlba::irlba(t(assay(mat2)), 30, 30,
                    center = rowMeans(assay(mat2)))
svdback = svd
svd$u = svdback$v
svd$v = svdback$u
rm(svdback)
```

Plot inspection

```{r}
library(viridisLite)

plotdata = as.data.frame(svd$v)
plotdata$colSm    = colSm
plotdata$ploidy   = mat2$ploidy
plotdata$Sample   = as.character(mat2$Sample)

ggplot(
  data = plotdata,
  aes(x = V1,
      y = V2)) +
  geom_point(aes(
    # col = log10(colSm)
    col = ploidy
    ),
    size = 0.1) +
  # scale_color_gradientn(colours = viridis(256, option = "D")) +
  facet_grid(rows = vars(Sample)) +
  geom_rug()
```

In above plot, higher dimensions appear to be random noise.

```{r}
dmax = 10
x = svd$v[, 1:dmax] %*% diag(svd$d[1:dmax])
rownames(x) = colnames(mat2)
colnames(x) = paste0("LSI", seq(1, ncol(x)))
```

```{r}
table(proj1$cellNames %in% rownames(x))
proj1@reducedDims[[
  "myLSI"
  ]] <-
  SimpleList(matDR = x[proj1$cellNames, ],
             params = list(),
             date = Sys.time(),
             scaleDims = NA, 
             corToDepth = NA)
```

# Remove outlier barcodes

Outliers are apparent in PC1 & PC2

```{r}
gc()
distmat = as.matrix(dist(svd$v[, 1:2] %*% diag(svd$d[1:2])))
distmat = distmat^2
distmat = apply(
  distmat,
  1,
  function (x) {
    sum(head(sort(x), 5 + 1)) # distance to itself is 0
  })
plot(log10(sort(distmat)))
outlier = (distmat > 10^-2)
table(outlier) # 14516    99
table(outlier, as.character(mat2$Sample))
# outlier m154207 m154211 m167108 m167203 m168101
#   FALSE    2891    3574    3001    2104    2946
#   TRUE        2       5      57       7      28
notoutlier = colnames(mat2)[! outlier]
rm(distmat)

ggplot(
  data = plotdata,
  aes(x = V1,
      y = V2)) +
  geom_point(aes(col = outlier),
             size = 0.1) +
  facet_grid(rows = vars(Sample)) +
  geom_rug()
```

```{r}
proj1b = proj1[which(proj1$cellNames %in% notoutlier), ]
proj1b@reducedDims[["myLSI"]]$matDR =
  (proj1b@reducedDims[["myLSI"]]$matDR)[proj1b$cellNames, ]
```

```{r}
rm(mat, mat2, svd, x)
```

# Clustering (1st round)

```{r}
proj1b <- addClusters(
    input = proj1b,
    reducedDims = "myLSI",
    method = "Seurat",
    algorithm = 4, # Leiden
    name = "Clusters",
    force = TRUE
)
table(proj1b$Sample, proj1b$Clusters)
```

Check cell types

```{r}
markerGenes = c(
  "Alb", "Apoc3", "Cyp2e1", "Cyp2f4",
  "Epcam", "Krt19",
  "Dcn", "Des", "Lrat",
  "Stab2",
  "Csf3r",
  "Elane", "Ly6g6c", "Ly6g6d", "Ly6g6e", "Mpo",
  "Csf1r",
  "Lyz2",
  "Marco",
  "Ebf1",
  "Cd3g", "Gata3",
  "Gzma", "Prf1"
)
```

```{r}
p <- plotBrowserTrack(
    ArchRProj = proj1b, 
    groupBy = "Clusters",
    geneSymbol = markerGenes, 
    upstream = 50000,
    downstream = 50000
)

plotPDF(plotList = p,
    name = "Plot-Tracks-Marker-Genes-Clusters.pdf", 
    ArchRProj = proj1b, 
    addDOC = FALSE, width = 5, height = 5)
```

```{r}
proj1b$Clusterscoarse = NA
proj1b$Clusterscoarse[
  proj1b$Clusters %in% paste0("C", 1:8)] = "hepatocyte"
proj1b$Clusterscoarse[
  proj1b$Clusters %in% paste0("C", c(9:10, 17:20))] = "WBC"
proj1b$Clusterscoarse[
  proj1b$Clusters %in% paste0("C", 13:16)] = "endothelial cell"
proj1b$Clusterscoarse[
  proj1b$Clusters %in% paste0("C", 11:12)] = "stellate"

ClusterscoarseLabels =
  unique(proj1b$Clusterscoarse)
```

# Loading and QC of tile matrix (2nd round)

Load tile matrix

```{r}
mat =
  getMatrixFromProject(
    proj1b,
    useMatrix = "TileMatrix",
    binarize = TRUE)
```

QC

```{r}
colSm <- Matrix::colSums(assay(mat))
rowSm <- Matrix::rowSums(assay(mat))

# Clean up zero/singleton rows
idx <- which(rowSm > 1)
mat <- mat[idx, ]
rowSm <- rowSm[idx]

# Exclude a small number of ~100% inserted windows
idx <- which(rowSm/ncol(mat) < 0.85)
mat <- mat[idx, ]
rowSm <- rowSm[idx]
```

# LSI (2nd round)

Smaller submatrix for faster computation

```{r}
# variance after IDF weighting
adjvar = log2(ncol(mat)/rowSm)^2 *
  (rowSm/ncol(mat)) *
  (1 - rowSm/ncol(mat))

idx = which(adjvar > adjvarmin)
mat2 = mat[idx, ]
rowSm2 = rowSm[idx]
```

Library size correct to unit vector

```{r}
assay(mat2) = assay(mat2) %*%
  Diagonal(x = 1/sqrt(colSums(assay(mat2)^2)))
```

Sample batch correct mat2

```{r}
for (cl in ClusterscoarseLabels) {
  print(cl)
  target = rowMeans(
    assay(mat2)[, which(mat2$Clusterscoarse == cl),
                drop = FALSE])
  
  for (s in sampleLabels) {
    print(s)
    x = assay(mat2) %*%
      Diagonal(x = as.numeric(mat2$Clusterscoarse == cl &
                                mat2$Sample == s))
    y = assay(mat2) %*%
      Diagonal(x = as.numeric(mat2$Clusterscoarse != cl |
                                mat2$Sample != s))
    rowMeansx = rowSums(x) / sum(mat2$Clusterscoarse == cl &
                                   mat2$Sample == s)
    correct = target / rowMeansx
    correct[rowMeansx == 0] = 1
    correctcap =
      quantile(correct, 0.75, na.rm = TRUE)^2 /
      quantile(correct, 0.5, na.rm = TRUE)
    correct = pmin(correct, correctcap)
    x = Diagonal(x = correct) %*% x

    assay(mat2) = x + y
    rm(x, y, rowMeansx, correct, correctcap)
  }
}
```

IDF (inverse document frequency)
rowSm2 is calculated from original binary matrix

```{r}
assay(mat2) = Diagonal(x = log2(ncol(mat2) / rowSm2)) %*%
  assay(mat2)
```

Row centralization is mandatory!
[Skillicorn, Understanding complex datasets. p.51]

```{r}
gc()
set.seed(1)

# irlba only provides column centralization, not for rows
svd <- irlba::irlba(t(assay(mat2)), 30, 30,
                    center = rowMeans(assay(mat2)))
svdback = svd
svd$u = svdback$v
svd$v = svdback$u
rm(svdback)
```

Plot inspection

```{r}
library(viridisLite)

plotdata = as.data.frame(svd$v)
plotdata$colSm          = colSm
plotdata$ploidy         = mat2$ploidy
plotdata$Clusterscoarse = mat2$Clusterscoarse
plotdata$Sample         = as.character(mat2$Sample)

ggplot(
  data = plotdata,
  aes(x = V1,
      y = V2)) +
  geom_point(aes(
    # col = log10(colSm)
    col = Clusterscoarse
    # col = ploidy
    ),
    size = 0.1) +
  # scale_color_gradientn(colours = viridis(256, option = "D")) +
  facet_grid(rows = vars(Sample)) +
  geom_rug()
```

In above plot, higher dimensions appear to be random noise.

```{r}
dmax = 8
x = svd$v[, 1:dmax] %*% diag(svd$d[1:dmax])
rownames(x) = colnames(mat2)
colnames(x) = paste0("LSI", seq(1, ncol(x)))
```

```{r}
table(proj1b$cellNames %in% rownames(x))
proj1b@reducedDims[[
  "myLSI2"
  ]] <-
  SimpleList(matDR = x[proj1b$cellNames, ],
             params = list(),
             date = Sys.time(),
             scaleDims = NA, 
             corToDepth = NA)
```

```{r}
rm(mat, mat2, svd, x)
```

# Clustering (2nd round)

```{r}
proj1b <- addClusters(
    input = proj1b,
    reducedDims = "myLSI2",
    method = "Seurat",
    algorithm = 4, # Leiden
    name = "Clusters2",
    force = TRUE
)

proj2 = proj1b
```

```{r}
table(proj2$Sample, proj2$Clusters)
table(proj2$Sample, proj2$Clusters2)
```

Check cell types

```{r}
p <- plotBrowserTrack(
    ArchRProj = proj2, 
    groupBy = "Clusters2",
    geneSymbol = markerGenes, 
    upstream = 50000,
    downstream = 50000
)

plotPDF(plotList = p,
    name = "Plot-Tracks-Marker-Genes-Clusters2.pdf", 
    ArchRProj = proj2, 
    addDOC = FALSE, width = 5, height = 5)
```

Clusters2
1-10  hepatocyte
11    B-cell, T-cell, NK cell
12    non-inflammatory macrophage, neutrophil
13    inflammatory macrophage, neutrophil
14-15 stellate
16-19 endothelial cell

```{r}
plotdata = as.data.frame(proj2@reducedDims[[
  "myLSI2"
  ]]$matDR)
all(proj2$cellNames == rownames(proj2@reducedDims[[
  "myLSI2"
  ]]$matDR))
plotdata$celltype = proj2$Clusters2
plotdata$Sample   = proj2$Sample

ggplot(
  data = plotdata,
  aes(x = LSI1,
      y = LSI2)) +
  geom_point(aes(
    col = celltype),
    size = 0.1) +
  facet_grid(rows = vars(Sample)) +
  geom_rug()
```

# Save result

```{r}
# saveArchRProject(
#   ArchRProj = proj2,
#   outputDirectory = "Save-Proj2",
#   dropCells = TRUE
# )

proj2 = loadArchRProject("Save-Proj2/")
```
