---
title: "LSI and clustering of peak matrix"
output: html_notebook
---

# Compute and save peak matrix

## Make pseudo-bulk replicates

Don't run this in parallel.

```{r}
proj2 = loadArchRProject("Save-Proj2/")

proj4 <- addGroupCoverages(
  ArchRProj = proj2,
  groupBy = "Clusters2")
```

## Call peaks w/ Macs2

```{r}
pathToMacs2 <- "/usr/local/anaconda2/bin/macs2"
```

Below took 282.13 mins

```{r}
Sys.setenv(HDF5_USE_FILE_LOCKING = "FALSE")
proj4 <- addReproduciblePeakSet(
    ArchRProj = proj4, 
    groupBy = "Clusters2",
    pathToMacs2 = pathToMacs2,
    # genomeSize = 2743300000 # rn6
    genomeSize = 2626580772 # rn7
)

getPeakSet(proj4)
```

## Add peak matrix

```{r}
# saveArchRProject(
#   ArchRProj = proj4,
#   outputDirectory = "Save-Proj4",
#   load = FALSE)
proj4 = loadArchRProject("Save-Proj4/")

# # This needs executed only once
# addPeakMatrix(proj4)

getAvailableMatrices(proj4)
```

# Loading and QC of peak matrix (1st round)

Load peak matrix

```{r}
proj5 = loadArchRProject("Save-Proj4/")
mat =
  getMatrixFromProject(
    proj5,
    useMatrix = "PeakMatrix",
    binarize = TRUE)
```

QC

```{r}
# colSm will not be updated. To be used later in QC plot.
colSm <- Matrix::colSums(assay(mat))

rowSm <- Matrix::rowSums(assay(mat))

# Clean up zero/singleton rows
# Actually redundant, because later filtered by adjvar
idx <- which(rowSm > 1)
mat <- mat[idx, ]
rowSm <- rowSm[idx]

# Exclude a small number of ~100% inserted windows
# Actually redundant, because later filtered by adjvar
idx <- which(rowSm/ncol(mat) < 0.85)
mat <- mat[idx, ]
rowSm <- rowSm[idx]
```

# LSI (1st round)

Smaller submatrix for faster computation

```{r}
adjvarmin = 0.1

# variance after IDF weighting
adjvar = log2(ncol(mat)/rowSm)^2 *
  (rowSm/ncol(mat)) *
  (1 - rowSm/ncol(mat))

qplot(
  x = rowSm/ncol(mat),
  y = adjvar,
  geom = "bin2d")
quantile(adjvar, seq(0, 1, 0.1))

idx = which(adjvar > adjvarmin)
mat2 = mat[idx, ]
rowSm2 = rowSm[idx]
```

Library size correct to unit vector

```{r}
assay(mat2) = assay(mat2) %*%
  Diagonal(x = 1/sqrt(colSums(assay(mat2)^2)))
```

Sample batch correct mat2

```{r}
target = rowMeans(assay(mat2))

sampleLabels = unique(as.character(mat$Sample))
for (s in sampleLabels) {
  print(s)
  x = assay(mat2) %*% Diagonal(x = as.numeric(mat2$Sample == s))
  y = assay(mat2) %*% Diagonal(x = as.numeric(mat2$Sample != s))
  rowMeansx = rowSums(x) / sum(mat2$Sample == s)
  correct = target / rowMeansx
  correct[rowMeansx == 0] = 1
  correctcap =
    quantile(correct, 0.75, na.rm = TRUE)^2 /
    quantile(correct, 0.5, na.rm = TRUE)
  correct = pmin(correct, correctcap)
  x = Diagonal(x = correct) %*% x

  assay(mat2) = x + y
  rm(x, y, rowMeansx, correct, correctcap)
}
```

IDF (inverse document frequency)
rowSm2 is calculated from original binary matrix

```{r}
assay(mat2) = Diagonal(x = log2(ncol(mat2) / rowSm2)) %*%
  assay(mat2)
```

Row centralization is mandatory!
[Skillicorn, Understanding complex datasets. p.51]

```{r}
gc()
set.seed(1)

# irlba only provides column centralization, not for rows
svd <- irlba::irlba(t(assay(mat2)), 30, 30,
                    center = rowMeans(assay(mat2)))
svdback = svd
svd$u = svdback$v
svd$v = svdback$u
rm(svdback)
```

Plot inspection

```{r}
library(viridisLite)

plotdata = as.data.frame(svd$v)
plotdata$colSm    = colSm
plotdata$ploidy   = mat2$ploidy
plotdata$Sample   = as.character(mat2$Sample)

ggplot(
  data = plotdata,
  aes(x = V1,
      y = V2)) +
  geom_point(aes(
    # col = log10(colSm)
    col = ploidy
    ),
    size = 0.1) +
  # scale_color_gradientn(colours = viridis(256, option = "D")) +
  facet_grid(rows = vars(Sample)) +
  geom_rug()
```

In above plot, higher dimensions appear to be random noise.

```{r}
dmax = 10
x = svd$v[, 1:dmax] %*% diag(svd$d[1:dmax])
rownames(x) = colnames(mat2)
colnames(x) = paste0("LSI", seq(1, ncol(x)))
```

```{r}
table(proj5$cellNames %in% rownames(x))
proj5@reducedDims[[
  "myLSI3"
  ]] <-
  SimpleList(matDR = x[proj5$cellNames, ],
             params = list(),
             date = Sys.time(),
             scaleDims = NA, 
             corToDepth = NA)
```

# Remove outlier barcodes

Outliers are apparent in PC1 & PC2

```{r}
gc()
distmat = as.matrix(dist(svd$v[, 1:2] %*% diag(svd$d[1:2])))
distmat = distmat^2
distmat = apply(
  distmat,
  1,
  function (x) {
    sum(head(sort(x), 5 + 1)) # distance to itself is 0
  })
plot(log10(sort(distmat)))
outlier = (distmat > 10^-2)
table(outlier) # 14486    30
table(outlier, as.character(mat2$Sample))
# outlier m154207 m154211 m167108 m167203 m168101
#   FALSE    2889    3572    2982    2103    2940
#   TRUE        2       2      19       1       6
notoutlier = colnames(mat2)[! outlier]
rm(distmat)

ggplot(
  data = plotdata,
  aes(x = V1,
      y = V2)) +
  geom_point(aes(col = outlier),
             size = 0.1) +
  facet_grid(rows = vars(Sample)) +
  geom_rug()
```

```{r}
proj5b = proj5[which(proj5$cellNames %in% notoutlier), ]
proj5b@reducedDims[["myLSI3"]]$matDR =
  (proj5b@reducedDims[["myLSI3"]]$matDR)[proj5b$cellNames, ]
```

```{r}
rm(mat, mat2, svd, x)
```

# Clustering (1st round)

```{r}
proj5b <- addClusters(
    input = proj5b,
    reducedDims = "myLSI3",
    method = "Seurat",
    algorithm = 4, # Leiden
    name = "Clusters3",
    force = TRUE
)
table(proj5b$Sample, proj5b$Clusters3)
```

Check cell types

```{r}
markerGenes = c(
  "Alb", "Apoc3", "Cyp2e1", "Cyp2f4",
  "Epcam", "Krt19",
  "Dcn", "Des", "Lrat",
  "Stab2",
  "Csf3r",
  "Elane", "Ly6g6c", "Ly6g6d", "Ly6g6e", "Mpo",
  "Csf1r",
  "Lyz2",
  "Marco",
  "Ebf1",
  "Cd3g", "Gata3",
  "Gzma", "Prf1"
)
```

```{r}
p <- plotBrowserTrack(
    ArchRProj = proj5b, 
    groupBy = "Clusters3",
    geneSymbol = markerGenes, 
    upstream = 50000,
    downstream = 50000
)

plotPDF(plotList = p,
    name = "Plot-Tracks-Marker-Genes-Clusters3.pdf", 
    ArchRProj = proj5b, 
    addDOC = FALSE, width = 5, height = 5)
```

```{r}
proj5b$Clusterscoarse = NA
proj5b$Clusterscoarse[
  proj5b$Clusters3 %in% paste0("C", 5:12)] = "hepatocyte"
proj5b$Clusterscoarse[
  proj5b$Clusters3 %in% paste0("C", c(1:4, 18:19))] = "WBC"
proj5b$Clusterscoarse[
  proj5b$Clusters3 %in% paste0("C", 15:17)] = "endothelial cell"
proj5b$Clusterscoarse[
  proj5b$Clusters3 %in% paste0("C", 13:14)] = "stellate"

ClusterscoarseLabels =
  unique(proj5b$Clusterscoarse)
```

# Loading and QC of peak matrix (2nd round)

Load peak matrix

```{r}
mat =
  getMatrixFromProject(
    proj5b,
    useMatrix = "PeakMatrix",
    binarize = TRUE)
```

QC

```{r}
colSm <- Matrix::colSums(assay(mat))
rowSm <- Matrix::rowSums(assay(mat))

# Clean up zero/singleton rows
idx <- which(rowSm > 1)
mat <- mat[idx, ]
rowSm <- rowSm[idx]

# Exclude a small number of ~100% inserted windows
idx <- which(rowSm/ncol(mat) < 0.85)
mat <- mat[idx, ]
rowSm <- rowSm[idx]
```

# LSI (2nd round)

Smaller submatrix for faster computation

```{r}
# variance after IDF weighting
adjvar = log2(ncol(mat)/rowSm)^2 *
  (rowSm/ncol(mat)) *
  (1 - rowSm/ncol(mat))

idx = which(adjvar > adjvarmin)
mat2 = mat[idx, ]
rowSm2 = rowSm[idx]
```

Library size correct to unit vector

```{r}
assay(mat2) = assay(mat2) %*%
  Diagonal(x = 1/sqrt(colSums(assay(mat2)^2)))
```

Sample batch correct mat2

```{r}
for (cl in ClusterscoarseLabels) {
  print(cl)
  target = rowMeans(
    assay(mat2)[, which(mat2$Clusterscoarse == cl),
                drop = FALSE])
  
  for (s in sampleLabels) {
    print(s)
    x = assay(mat2) %*%
      Diagonal(x = as.numeric(mat2$Clusterscoarse == cl &
                                mat2$Sample == s))
    y = assay(mat2) %*%
      Diagonal(x = as.numeric(mat2$Clusterscoarse != cl |
                                mat2$Sample != s))
    rowMeansx = rowSums(x) / sum(mat2$Clusterscoarse == cl &
                                   mat2$Sample == s)
    correct = target / rowMeansx
    correct[rowMeansx == 0] = 1
    correctcap =
      quantile(correct, 0.75, na.rm = TRUE)^2 /
      quantile(correct, 0.5, na.rm = TRUE)
    correct = pmin(correct, correctcap)
    x = Diagonal(x = correct) %*% x

    assay(mat2) = x + y
    rm(x, y, rowMeansx, correct, correctcap)
  }
}
```

IDF (inverse document frequency)
rowSm2 is calculated from original binary matrix

```{r}
assay(mat2) = Diagonal(x = log2(ncol(mat2) / rowSm2)) %*%
  assay(mat2)
```

Row centralization is mandatory!
[Skillicorn, Understanding complex datasets. p.51]

```{r}
gc()
set.seed(1)

# irlba only provides column centralization, not for rows
svd <- irlba::irlba(t(assay(mat2)), 30, 30,
                    center = rowMeans(assay(mat2)))
svdback = svd
svd$u = svdback$v
svd$v = svdback$u
rm(svdback)
```

Plot inspection

```{r}
library(viridisLite)

plotdata = as.data.frame(svd$v)
plotdata$colSm          = colSm
plotdata$ploidy         = mat2$ploidy
plotdata$Clusterscoarse = mat2$Clusterscoarse
plotdata$Sample         = as.character(mat2$Sample)

ggplot(
  data = plotdata,
  aes(x = V1,
      y = V2)) +
  geom_point(aes(
    # col = log10(colSm)
    col = Clusterscoarse
    # col = ploidy
    ),
    size = 0.1) +
  # scale_color_gradientn(colours = viridis(256, option = "D")) +
  facet_grid(rows = vars(Sample)) +
  geom_rug()
```

In above plot, higher dimensions appear to be random noise.

```{r}
dmax = 8
x = svd$v[, 1:dmax] %*% diag(svd$d[1:dmax])
rownames(x) = colnames(mat2)
colnames(x) = paste0("LSI", seq(1, ncol(x)))
```

```{r}
table(proj5b$cellNames %in% rownames(x))
proj5b@reducedDims[[
  "myLSI4"
]] <-
  SimpleList(matDR = x[proj5b$cellNames, ],
             params = list(),
             date = Sys.time(),
             scaleDims = NA, 
             corToDepth = NA)
```

```{r}
rm(mat, mat2, svd, x)
```

# Clustering (2nd round)

```{r}
proj5b <- addClusters(
    input = proj5b,
    reducedDims = "myLSI4",
    method = "Seurat",
    algorithm = 4, # Leiden
    name = "Clusters4",
    force = TRUE
)

proj6 = proj5b
```

```{r}
table(proj6$Sample, proj6$Clusters3)
table(proj6$Sample, proj6$Clusters4)
```

Check cell types

```{r}
p <- plotBrowserTrack(
    ArchRProj = proj6, 
    groupBy = "Clusters4",
    geneSymbol = markerGenes, 
    upstream = 50000,
    downstream = 50000
)

plotPDF(plotList = p,
    name = "Plot-Tracks-Marker-Genes-Clusters4.pdf", 
    ArchRProj = proj6, 
    addDOC = FALSE, width = 5, height = 5)
```

Clusters4
1-11  hepatocyte (6 cholangiocyte)
13    T-cell, NK cell
14    B-cell
12    non-inflammatory macrophage, neutrophil
15    inflammatory macrophage, neutrophil
16-17 stellate
18-22 endothelial cell

```{r}
plotdata = as.data.frame(proj6@reducedDims[[
  "myLSI4"
  ]]$matDR)
all(proj6$cellNames == rownames(proj6@reducedDims[[
  "myLSI4"
  ]]$matDR))
plotdata$celltype = proj6$Clusters4
plotdata$Sample   = proj6$Sample

ggplot(
  data = plotdata,
  aes(x = LSI1,
      y = LSI2)) +
  geom_point(aes(
    col = celltype),
    size = 0.1) +
  facet_grid(rows = vars(Sample)) +
  geom_rug()
```

# Clustering with taylored parameter

```{r}
proj6 <- addClusters(
    input = proj6,
    reducedDims = "myLSI4",
    method = "Seurat",
    algorithm = 4, # Leiden
    resolution = 0.6,
    name = "Clusters4res06",
    force = TRUE
)
```

Clusters4res05
1-6   hepatocyte
7     B-cell, T-cell, NK cell
8     non-inflammatory macrophage, neutrophil
9     inflammatory macrophage, neutrophil
12-13 stellate
10-11 endothelial cell

Clusters4res055
1-7   hepatocyte (5 cholangiocyte)
13    T-cell, NK cell
12    B-cell
10    non-inflammatory macrophage, neutrophil
11    inflammatory macrophage, neutrophil
8-9   stellate
14-16 endothelial cell

** Clusters4res06 **
1-7   hepatocyte
14    T-cell, NK cell
15    B-cell
10    non-inflammatory macrophage, neutrophil
16    inflammatory macrophage, neutrophil
8-9   stellate
11-13 endothelial cell

Clusters4res065
1-9   hepatocyte
15    T-cell, NK cell
14    B-cell
12    non-inflammatory macrophage, neutrophil
13    inflammatory macrophage, neutrophil
10-11 stellate
16-18 endothelial cell

For markers
- Lyz2 for inflammatory macrophage
- Marco for non-inflammatory macrophage
Clusters4res06 was more appropriate than Clusters4res055
Clusters4res06 was more appropriate than Clusters4res065

```{r}
p <- plotBrowserTrack(
    ArchRProj = proj6, 
    groupBy = "Clusters4res06",
    geneSymbol = markerGenes, 
    upstream = 50000,
    downstream = 50000
)

plotPDF(plotList = p,
    name = "Plot-Tracks-Marker-Genes-Clusters4res06.pdf", 
    ArchRProj = proj6, 
    addDOC = FALSE, width = 5, height = 5)
```

```{r}
g = proj6$Clusters4res06
g[g %in% paste0("C", 1:7)] = "hepatocyte"
g[g %in% paste0("C", c(10, 14:16))] = "WBC"
g[g %in% paste0("C", 8:9)] = "stellate"
g[g %in% paste0("C", 11:13)] = "endothelial"
proj6$Clusterscoarse2 = g

g = proj6$Clusters4res06
g[g %in% paste0("C", 1:7)] = "hepato"
g[g %in% paste0("C", c(10, 16))] = "myelo"
g[g %in% paste0("C", 14)] = "T_NK"
g[g %in% paste0("C", 15)] = "B"
g[g %in% paste0("C", 8:9)] = "stellate"
g[g %in% paste0("C", 11:13)] = "endothelial"
proj6$Clusters4BTmyelo = g
```

## Uniform Manifold Approximation and Projection (UMAP)

```{r}
proj6 <- addUMAP(
    ArchRProj = proj6, 
    reducedDims = "myLSI4",
    name = "UMAP", 
    metric = "cosine"
)

p1 <- plotEmbedding(
  ArchRProj = proj6,
  colorBy = "cellColData",
  name = "Sample",
  embedding = "UMAP")

p2 <- plotEmbedding(
  ArchRProj = proj6,
  colorBy = "cellColData",
  name = "Clusters4res06",
  embedding = "UMAP")

ggAlignPlots(p1, p2, type = "h")

plotPDF(
  p1,p2,
  name = "Plot-UMAP-Sample-Clusters.pdf",
  ArchRProj = proj6,
  addDOC = FALSE,
  width = 5,
  height = 5)
```

## Make pseudo-bulk replicates

Don't run this in parallel.

```{r}
proj6 <- addGroupCoverages(
  ArchRProj = proj6,
  groupBy = "Clusters4res06")
```

# Save result

```{r}
# saveArchRProject(
#   ArchRProj = proj6,
#   outputDirectory = "Save-Proj6",
#   dropCells = TRUE
# )

proj6 = loadArchRProject("Save-Proj6/")
```
