---
title: "MotifMatrix"
output: html_notebook
---

# Add MotifMatrix

```{r}
proj6 = loadArchRProject("Save-Proj6/")
```

The following computation took 16 mins

```{r}
proj6 <- addMotifAnnotations(
  ArchRProj = proj6,
  motifSet = "cisbp",
  species = "Mus musculus",
  name = "Motif")

proj6@peakAnnotation@listData$Motif$motifSummary
```

The following computation took 54 mins

```{r}
proj6 <- addBgdPeaks(proj6)

proj6 <- addDeviationsMatrix(
  ArchRProj = proj6, 
  peakAnnotation = "Motif",
  force = TRUE)
```

```{r}
mat = getMatrixFromProject(proj6, "MotifMatrix")
mat@assays@data@listData =
  mat@assays@data@listData["z"] # drop deviations assay

idx = which(rowSums(is.na(assay(mat))) == 0)
mat = mat[idx, ]
```

Convert rownames to rat gene symbol

```{r}
dimnames(mat)[[1]] = 
  sub("_.*", "", dimnames(mat)[[1]])
# Tcfap2a_1 -> Tcfap2a

attach("~/human/publicrat/homologs/mouse_to_rat_homologs.Rda")
x = mouse_to_rat_homologs$RGD.symbol[
  match(
    rownames(mat),
    mouse_to_rat_homologs$NCBI.gene..formerly.Entrezgene..accession)]
dimnames(mat)[[1]] = x
idx = which((! is.na(rownames(mat))) & (rownames(mat) != ""))
mat = mat[idx, ]

idx = which(! duplicated(rownames(mat)))
mat = mat[idx, ]
```

# QC

Library size correct (qn)

```{r}
assay(mat) = as.matrix(assay(mat))
x = preprocessCore::normalize.quantiles(assay(mat))
assay(mat, withDimnames = FALSE) = x
```

Row centralize

```{r}
assay(mat) = assay(mat) - rowMeans(assay(mat))
```

Compute and plot SVD without sample batch correction

```{r}
set.seed(1)
svd = irlba::irlba(assay(mat), nv = 20)
plot(svd$d)
ggplot(
  data = as.data.frame(svd$v)) +
  geom_point(aes(x = V1, y = V2), size = 0.3) +
  facet_grid(
    cols = vars(mat$Clusters4BTmyelo),
    rows = vars(as.character(mat$Sample)))
```

The first two components capture
A. sample batch effect, and
B. difference between major cell types.
In the following, we analyze the effect of diet intervention by comparing samples.
Specifically, we compare nuclei of same cell type between different sample.
Thus, we prefer to eliminate A, hoping that the removal of B does not harm.

```{r}
assay(mat) = assay(mat) -
  svd$u[, 1:2] %*% diag(svd$d[1:2]) %*% t(svd$v[, 1:2])
```

# Save result

```{r}
saveRDS(mat, file = "matMotifSamplebatchcorrected.rds")
```
