---
title: "GSEA"
output: html_notebook
---

# Load GeneScoreMatrix

```{r}
proj6 = loadArchRProject("Save-Proj6/")
mat = readRDS("matGeneScoreSamplebatchcorrected.rds")
```

# Define gene sets for GSEA

```{r}
library(msigdbr)

gslistfrombaderlab = function (x) {
  x = strsplit(x, "\t")
  gslist = lapply(x, function (x) { x[-c(1:2)] })
  names(gslist) = unlist(lapply(x, function (x) { x[1] }))
  return(gslist)
}
gslistfrommsigdbr = function (x) {
  x = x %>%
    dplyr::mutate(name = paste0(gs_name, "%%", gs_id)) %>%
    dplyr::select(name, gene_symbol) %>%
    dplyr::group_by(name) %>%
    dplyr::summarize(gs = list(gene_symbol))
  gslist = x$gs
  names(gslist) = x$name
  return(gslist)
}

gslists = list()
x = readLines("~/human/publicrat/baderlab_EM_Genesets/Rat_Human_NetPath_June_01_2021_symbol.gmt")
gslists = c(gslists,
            list(NetPath = gslistfrombaderlab(x)))
x = readLines("~/human/publicrat/baderlab_EM_Genesets/Rat_Human_Panther_June_01_2021_symbol.gmt")
gslists = c(gslists,
            list(Panther = gslistfrombaderlab(x)))
x = msigdbr(species = "Rattus norvegicus", category = "H")
gslists = c(gslists,
            list(H = gslistfrommsigdbr(x)))
x = msigdbr(species = "Rattus norvegicus", category = "C2", subcategory = "CP:KEGG")
gslists = c(gslists,
            list(KEGG = gslistfrommsigdbr(x)))
x = msigdbr(species = "Rattus norvegicus", category = "C2", subcategory = "CP:PID")
gslists = c(gslists,
            list(PID = gslistfrommsigdbr(x)))
x = msigdbr(species = "Rattus norvegicus", category = "C2", subcategory = "CP:REACTOME")
gslists = c(gslists,
            list(REACTOME = gslistfrommsigdbr(x)))
x = msigdbr(species = "Rattus norvegicus", category = "C2", subcategory = "CP:WIKIPATHWAYS")
gslists = c(gslists,
            list(WIKIPATHWAYS = gslistfrommsigdbr(x)))
x1 = msigdbr(species = "Rattus norvegicus", category = "C3", subcategory = "TFT:GTRD")
x2 = msigdbr(species = "Rattus norvegicus", category = "C3", subcategory = "TFT:TFT_Legacy")
gslists = c(gslists,
            list(TFT = c(gslistfrommsigdbr(x1), gslistfrommsigdbr(x2))))
x = msigdbr(species = "Rattus norvegicus", category = "C5", subcategory = "GO:BP")
gslists = c(gslists,
            list(GOBP = gslistfrommsigdbr(x)))
rm(x, x1, x2)
```

# Define conditions to be compared

```{r}
g = proj6$Sample
g[g %in% c("m154211", "m167108")] = "sp"
proj6$spdietcombined = g

proj6$Clusters4BTmyelospdietcombined =
  paste0(proj6$Clusters4BTmyelo,
         "_",
         proj6$spdietcombined)
```

Compare nuclei of same cell type between different interventions
Choose tgt vs bgd

```{r}
gname = "Clusters4BTmyelospdietcombined"
g = getCellColData(proj6, gname, drop = TRUE)[
  match(colnames(mat),
        proj6$cellNames)]
ClustersLabels = unique(as.character(g))

tgt = "hepato_m154207"; bgd = "hepato_sp"
tgt = "hepato_m167203"; bgd = "hepato_sp"
tgt = "hepato_m168101"; bgd = "hepato_sp"

tgt = "stellate_m154207"; bgd = "stellate_sp"
tgt = "stellate_m167203"; bgd = "stellate_sp"
tgt = "stellate_m168101"; bgd = "stellate_sp"

tgt = "endothelial_m154207"; bgd = "endothelial_sp"
tgt = "endothelial_m167203"; bgd = "endothelial_sp"
tgt = "endothelial_m168101"; bgd = "endothelial_sp"

tgt = "myelo_m154207";  bgd = "myelo_sp"
tgt = "myelo_m167203";  bgd = "myelo_sp"
tgt = "myelo_m168101";  bgd = "myelo_sp"

tgt = "B_m154207";  bgd = "B_sp"
tgt = "B_m167203";  bgd = "B_sp"
tgt = "B_m168101";  bgd = "B_sp"

tgt = "T_NK_m154207";  bgd = "T_NK_sp"
tgt = "T_NK_m167203";  bgd = "T_NK_sp"
tgt = "T_NK_m168101";  bgd = "T_NK_sp"
```

Compare clusters

```{r}
gname = "Clusters4res06"
g = getCellColData(proj6, gname, drop = TRUE)[
  match(colnames(mat),
        proj6$cellNames)]
ClustersLabels = unique(as.character(g))
ClustersLabels = ClustersLabels[
  order(as.numeric(sub("^C", "", ClustersLabels)))]

tgt = "C16"; bgd = "C10" # inflammatory vs non-inflammatory macrophage
tgt =  "C8"; bgd =  "C9" # stellate
tgt = "C11"; bgd = "C12" # endothelial cell
```

# Run padog GSEA

```{r}
esetm = cbind(
  assay(mat)[, which(g == tgt)],
  assay(mat)[, which(g == bgd)])
group = c(rep("d", sum(g == tgt)), rep("c", sum(g == bgd)))

library(PADOG)
dseed = 1
result = list()
for (i in 1:length(gslists)) {
  print(names(gslists)[i])
  myr = padog(
    esetm = esetm,
    group = group,
    gslist = gslists[[i]],
    organism = "rno",
    Nmin = 5,
    NI = max(10000, length(gslists[[i]]) * 20),
    parallel = TRUE,
    ncr = 16,
    dseed = dseed)
  myr$FDRmeanAbsT = p.adjust(myr$PmeanAbsT, method = "BH")
  myr$FDRpadog    = p.adjust(myr$Ppadog, method = "BH")
  result = c(result, list(myr))
}
names(result) = names(gslists)

saveRDS(result, file = paste0(
  "padog/",
  paste0(c(gname, tgt, bgd, dseed, "rds"), collapse = ".")))
```

# Inspect enriched gene sets

When there are too many gene sets attaining FDRpadog < 0.05,
visualize and choose representatives by using Cytoscape:
1. Apps > EnrichmentMap. Load GMT file `foo.gmt`
2. Node Table. Load `foo.txt` and add column log10Ppadog
2. Style > Fill color. Column = log10Ppadog. Mapping Type = Continuous Mapping
3. Layout > yFiles Radial Layout

As above, define `gname`, `tgt`, `bgd`.
Generate files to be loaded to Cytoscape.

```{r}
dseed = 1
padog =
  readRDS(paste0(
    "padog/",
    # "padog.GeneScore/",
    # "padog.Motif/",
    paste0(c(gname, tgt, bgd, dseed, "rds"), collapse = ".")))

x = padog[c(1:7, 9)]
x = do.call(rbind, x)
padogFDR005 = x[x$FDRpadog < 0.05, ]
padogFDR005$log10Ppadog = log10(padogFDR005$Ppadog)
write.table(
  padogFDR005[, c("Name", "log10Ppadog")],
  file = "foo.txt",
  row.names = FALSE,
  col.names = TRUE,
  sep = "\t",
  quote = FALSE)

output = do.call(c, gslists)
output = output[rownames(padogFDR005)]
output = lapply(output, function (x) { paste0(x, collapse = "\t") })
x = names(output)
x = sub("[^.]*\\.", "", x, perl = TRUE)
output = paste(
  x,
  sub("%.*", "", x),
  unlist(output),
  sep = "\t")
cat(output, file = "foo.gmt", sep = "\n")
```

# Result: representative enriched gene sets

```{r}
gname = "Clusters4BTmyelospdietcombined"

tgt = "hepato_m154207"; bgd = "hepato_sp"
# dbname ="Panther"; gsname = "CHOLESTEROL BIOSYNTHESIS%PANTHER PATHWAY%P00014"
# dbname ="H"; gsname = "HALLMARK_XENOBIOTIC_METABOLISM%%M5934"
# dbname ="H"; gsname = "HALLMARK_BILE_ACID_METABOLISM%%M5948"
# dbname ="H"; gsname = "HALLMARK_CHOLESTEROL_HOMEOSTASIS%%M5892"
# dbname ="KEGG"; gsname = "KEGG_PRIMARY_BILE_ACID_BIOSYNTHESIS%%M3214"
# dbname ="KEGG"; gsname = "KEGG_STEROID_BIOSYNTHESIS%%M5872"
# dbname ="PID"; gsname = "PID_HNF3B_PATHWAY%%M106"
# dbname ="PID"; gsname = "PID_IL2_PI3K_PATHWAY%%M143"
dbname ="REACTOME"; gsname = "REACTOME_METABOLISM_OF_STEROIDS%%M27832"
dbname ="REACTOME"; gsname = "REACTOME_CHOLINE_CATABOLISM%%M27619"
# dbname ="REACTOME"; gsname = "REACTOME_BILE_ACID_AND_BILE_SALT_METABOLISM%%M499"
# dbname ="REACTOME"; gsname = "REACTOME_XENOBIOTICS%%M5372"
# dbname ="REACTOME"; gsname = "REACTOME_CHOLESTEROL_BIOSYNTHESIS%%M16227"
# dbname ="REACTOME"; gsname = "REACTOME_FATTY_ACID_METABOLISM%%M27854"
# dbname ="WIKIPATHWAYS"; gsname = "WP_CHOLESTEROL_BIOSYNTHESIS_PATHWAY%%M39608"
dbname ="WIKIPATHWAYS"; gsname = "WP_SIGNAL_TRANSDUCTION_THROUGH_IL1R%%M39803"
dbname ="GOBP"; gsname = "GOBP_ALPHA_AMINO_ACID_METABOLIC_PROCESS%%M10829"
dbname ="GOBP"; gsname = "GOBP_FATTY_ACID_METABOLIC_PROCESS%%M17829"
dbname ="GOBP"; gsname = "GOBP_LIPID_LOCALIZATION%%M15749"
# dbname ="GOBP"; gsname = "GOBP_LIPID_BIOSYNTHETIC_PROCESS%%M14504"
# dbname ="GOBP"; gsname = "GOBP_CHOLINE_CATABOLIC_PROCESS%%M23533"
# dbname ="GOBP"; gsname = "GOBP_STEROID_BIOSYNTHETIC_PROCESS%%M13381"
dbname ="GOBP"; gsname = "GOBP_RESPONSE_TO_DRUG%%M6886"
# dbname ="GOBP"; gsname = "GOBP_REGULATION_OF_LIPID_LOCALIZATION%%M25391"
# dbname ="GOBP"; gsname = "GOBP_REGULATION_OF_LIPID_TRANSPORT%%M13620"
# dbname ="GOBP"; gsname = "GOBP_BILE_ACID_BIOSYNTHETIC_PROCESS%%M12041"
tgt = "hepato_m167203"; bgd = "hepato_sp"
# dbname ="NetPath"; gsname = "IL1%NETPATH%IL1"  # up Il1r1 Il1r2 Nfkb1 relevant? NetPath reliable?
dbname ="Panther"; gsname = "APOPTOSIS SIGNALING PATHWAY%PANTHER PATHWAY%P00006"  # up
# dbname ="Panther"; gsname = "PLASMINOGEN ACTIVATING CASCADE%PANTHER PATHWAY%P00050"  # up
dbname ="H"; gsname = "HALLMARK_TNFA_SIGNALING_VIA_NFKB%%M5890"
# dbname ="H"; gsname = "HALLMARK_IL6_JAK_STAT3_SIGNALING%%M5897"
# dbname ="PID"; gsname = "PID_IL1_PATHWAY%%M110"
# dbname ="PID"; gsname = "PID_IL2_PI3K_PATHWAY%%M143"
dbname ="REACTOME"; gsname = "REACTOME_RESPONSE_TO_METAL_IONS%%M27541"
# dbname ="REACTOME"; gsname = "REACTOME_INTERLEUKIN_10_SIGNALING%%M27605"
dbname ="REACTOME"; gsname = "REACTOME_INTERLEUKIN_1_SIGNALING%%M27895"
# dbname ="REACTOME"; gsname = "REACTOME_CYTOKINE_SIGNALING_IN_IMMUNE_SYSTEM%%M1060"
# dbname ="REACTOME"; gsname = "REACTOME_SIGNALING_BY_INTERLEUKINS%%M874"
# dbname ="REACTOME"; gsname = "REACTOME_CYP2E1_REACTIONS%%M27130"
# dbname ="REACTOME"; gsname = "REACTOME_XENOBIOTICS%%M5372"
# dbname ="REACTOME"; gsname = "REACTOME_BILE_ACID_AND_BILE_SALT_METABOLISM%%M499"
dbname ="REACTOME"; gsname = "REACTOME_INFLAMMASOMES%%M1072"
# dbname ="WIKIPATHWAYS"; gsname = "WP_NUCLEAR_RECEPTORS_IN_LIPID_METABOLISM_AND_TOXICITY%%M39488"
dbname ="WIKIPATHWAYS"; gsname = "WP_SIGNAL_TRANSDUCTION_THROUGH_IL1R%%M39803"
# dbname ="TFT"; gsname = "NFKB_C%%M12240"
dbname ="GOBP"; gsname = "GOBP_CYTOKINE_PRODUCTION%%M13156"
dbname ="GOBP"; gsname = "GOBP_REGULATION_OF_CELL_CELL_ADHESION%%M6401"
# dbname ="GOBP"; gsname = "GOBP_CYTOKINE_MEDIATED_SIGNALING_PATHWAY%%M898"  # up
dbname ="GOBP"; gsname = "GOBP_INFLAMMATORY_RESPONSE%%M10617"
# dbname ="GOBP"; gsname = "GOBP_REGULATION_OF_CELL_ADHESION%%M8154"
dbname ="GOBP"; gsname = "GOBP_RESPONSE_TO_LIPID%%M14037"  # up&down
dbname ="GOBP"; gsname = "GOBP_RESPONSE_TO_DRUG%%M6886"
# dbname ="GOBP"; gsname = "GOBP_REGULATION_OF_INTERLEUKIN_1_MEDIATED_SIGNALING_PATHWAY%%M25511"
# dbname ="GOBP"; gsname = "GOBP_REGULATION_OF_LIPID_LOCALIZATION%%M25391"  # up&down
dbname ="GOBP"; gsname = "GOBP_RESPONSE_TO_CYTOKINE%%M15783"
dbname ="GOBP"; gsname = "GOBP_LIPID_LOCALIZATION%%M15749"
tgt = "hepato_m168101"; bgd = "hepato_sp"
# none

tgt = "stellate_m154207"; bgd = "stellate_sp"
# dbname ="KEGG"; gsname = "KEGG_LYSINE_DEGRADATION%%M13720"
# dbname ="KEGG"; gsname = "KEGG_STEROID_BIOSYNTHESIS%%M5872"   # rno00100 down
# dbname ="KEGG"; gsname = "KEGG_TERPENOID_BACKBONE_BIOSYNTHESIS%%M13465"
dbname ="REACTOME"; gsname = "REACTOME_CHOLESTEROL_BIOSYNTHESIS%%M16227"
# dbname ="WIKIPATHWAYS"; gsname = "WP_CHOLESTEROL_BIOSYNTHESIS_PATHWAY%%M39608"
tgt = "stellate_m167203"; bgd = "stellate_sp"
dbname ="GOBP"; gsname = "GOBP_ACTIN_FILAMENT_BASED_PROCESS%%M9082"  #up&down
dbname ="GOBP"; gsname = "GOBP_CELL_MIGRATION%%M40389"  #up&down
# dbname ="GOBP"; gsname = "GOBP_REGULATION_OF_CELLULAR_COMPONENT_MOVEMENT%%M15511"
# dbname ="GOBP"; gsname = "GOBP_ENDOTHELIAL_CELL_FATE_COMMITMENT%%M24225"
# dbname ="GOBP"; gsname = "GOBP_EPITHELIAL_CELL_FATE_COMMITMENT%%M16390"  # up&down
tgt = "stellate_m168101"; bgd = "stellate_sp"
# none

tgt = "endothelial_m154207"; bgd = "endothelial_sp"
dbname ="NetPath"; gsname = "IL2%NETPATH%IL2"
# dbname ="NetPath"; gsname = "IL4%NETPATH%IL4"
tgt = "endothelial_m167203"; bgd = "endothelial_sp"
# dbname ="NetPath"; gsname = "WNT%NETPATH%WNT"
# dbname ="NetPath"; gsname = "ALPHA6BETA4INTEGRIN%NETPATH%ALPHA6BETA4INTEGRIN"
dbname ="GOBP"; gsname = "GOBP_STEROID_METABOLIC_PROCESS%%M16758"  # down
tgt = "endothelial_m168101"; bgd = "endothelial_sp"
# none

tgt = "myelo_m154207";  bgd = "myelo_sp"
# dbname ="REACTOME"; gsname = "REACTOME_COMPLEMENT_CASCADE%%M19752"  # 0.04894015 irrelevant?
# dbname ="GOBP"; gsname = "GOBP_ALPHA_BETA_T_CELL_DIFFERENTIATION%%M10942"  # irrelevant?
dbname ="GOBP"; gsname = "GOBP_REGULATION_OF_IMMUNE_SYSTEM_PROCESS%%M5017"
tgt = "myelo_m167203";  bgd = "myelo_sp"
dbname ="GOBP"; gsname = "GOBP_CELLULAR_LIPID_METABOLIC_PROCESS%%M14398"  # up&down
tgt = "myelo_m168101";  bgd = "myelo_sp"
# none

tgt = "B_m154207";  bgd = "B_sp"
# none
tgt = "B_m167203";  bgd = "B_sp"
# none
tgt = "B_m168101";  bgd = "B_sp"
# none

tgt = "T_NK_m154207";  bgd = "T_NK_sp"
dbname ="GOBP"; gsname = "GOBP_FATTY_ACID_METABOLIC_PROCESS%%M17829"
dbname ="H"; gsname = "HALLMARK_XENOBIOTIC_METABOLISM%%M5934"
tgt = "T_NK_m167203";  bgd = "T_NK_sp"
dbname ="GOBP"; gsname = "GOBP_STEROID_METABOLIC_PROCESS%%M16758"
dbname ="GOBP"; gsname = "GOBP_ORGANIC_ACID_METABOLIC_PROCESS%%M14117"
dbname ="GOBP"; gsname = "GOBP_RESPONSE_TO_DRUG%%M6886"
tgt = "T_NK_m168101";  bgd = "T_NK_sp"
# none
```

Not significant but important

```{r}
dbname ="WIKIPATHWAYS"; gsname = "WP_NONALCOHOLIC_FATTY_LIVER_DISEASE%%M39806"
#
dbname ="KEGG"; gsname = "KEGG_CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION%%M9809"
#
dbname ="REACTOME"; gsname = "REACTOME_EXTRACELLULAR_MATRIX_ORGANIZATION%%M610"
dbname ="GOBP"; gsname = "GOBP_CONNECTIVE_TISSUE_REPLACEMENT%%M24786"
#
dbname ="KEGG"; gsname = "KEGG_TGF_BETA_SIGNALING_PATHWAY%%M2642"  # rno04350
dbname ="WIKIPATHWAYS"; gsname = "WP_TGFBETA_SIGNALING_PATHWAY%%M39432"
dbname ="Panther"; gsname = "TGF-BETA SIGNALING PATHWAY%PANTHER PATHWAY%P00052"
#
dbname ="WIKIPATHWAYS"; gsname = "WP_FERROPTOSIS%%M39768"
dbname ="REACTOME"; gsname = "REACTOME_PYROPTOSIS%%M41804"  # FDR >0.9
dbname ="REACTOME"; gsname = "REACTOME_INTRINSIC_PATHWAY_FOR_APOPTOSIS%%M7455"  # FDR >0.8
dbname ="REACTOME"; gsname = "REACTOME_SUPPRESSION_OF_APOPTOSIS%%M29815"  # FDR >0.7
dbname ="KEGG"; gsname = "KEGG_APOPTOSIS%%M8492"  # rno04210 FDR >0.5
dbname ="GOBP"; gsname = "GOBP_APOPTOTIC_PROCESS%%M34075"
```

```{r}
gname = "Clusters4res05"

tgt = "C9";  bgd = "C8"
dbname ="GOBP"; gsname = "GOBP_CELL_MIGRATION%%M40389"
dbname ="GOBP"; gsname = "GOBP_CELL_CELL_ADHESION%%M12087"
dbname ="GOBP"; gsname = "GOBP_RESPONSE_TO_WOUNDING%%M5634"  # up&down
dbname ="GOBP"; gsname = "GOBP_LEUKOCYTE_CELL_CELL_ADHESION%%M14645"
dbname ="GOBP"; gsname = "GOBP_LEUKOCYTE_MIGRATION%%M12269"  # up&down
tgt = "C14";  bgd = "C13"
dbname ="H"; gsname = "HALLMARK_TNFA_SIGNALING_VIA_NFKB%%M5890"  # up&down
dbname ="KEGG"; gsname = "KEGG_CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION%%M9809"  # mostly up
dbname ="KEGG"; gsname = "KEGG_LEUKOCYTE_TRANSENDOTHELIAL_MIGRATION%%M2164"  # up&down
dbname ="GOBP"; gsname = "GOBP_CELL_MIGRATION%%M40389"
dbname ="GOBP"; gsname = "GOBP_REGULATION_OF_CELL_ADHESION%%M8154"
dbname ="GOBP"; gsname = "GOBP_BLOOD_VESSEL_MORPHOGENESIS%%M13415"  # up&down
dbname ="GOBP"; gsname = "GOBP_POSITIVE_REGULATION_OF_REACTIVE_OXYGEN_SPECIES_METABOLIC_PROCESS%%M13580"  # up&down
dbname ="GOBP"; gsname = "GOBP_INFLAMMATORY_RESPONSE%%M10617"  # mostly up
dbname ="GOBP"; gsname = "GOBP_CYTOKINE_PRODUCTION%%M13156"
dbname ="GOBP"; gsname = "GOBP_RESPONSE_TO_WOUNDING%%M5634"  # up&down
```

For MotifMatrix

```{r}
# Motif  size >6
tgt = "hepato_m154207"; bgd = "hepato_sp"
tgt = "hepato_m167203"; bgd = "hepato_sp"
tgt = "hepato_m168101"; bgd = "hepato_sp"
dbname ="WIKIPATHWAYS"; gsname = "WP_IL2_SIGNALING_PATHWAY%%M39536"  # size 7
dbname ="WIKIPATHWAYS"; gsname = "WP_IL5_SIGNALING_PATHWAY%%M39392"  # size 8
dbname ="GOBP"; gsname = "GOBP_RECEPTOR_SIGNALING_PATHWAY_VIA_STAT%%M24785"  # size 11

tgt = "stellate_m154207"; bgd = "stellate_sp"
tgt = "stellate_m167203"; bgd = "stellate_sp"
dbname ="H"; gsname = "HALLMARK_TNFA_SIGNALING_VIA_NFKB%%M5890"  # size 39
tgt = "stellate_m168101"; bgd = "stellate_sp"

tgt = "endothelial_m154207"; bgd = "endothelial_sp"
tgt = "endothelial_m167203"; bgd = "endothelial_sp"
dbname ="GOBP"; gsname = "GOBP_RESPONSE_TO_REACTIVE_OXYGEN_SPECIES%%M16953"  # size 12
tgt = "endothelial_m168101"; bgd = "endothelial_sp"

tgt = "myelo_m154207";  bgd = "myelo_sp"
tgt = "myelo_m167203";  bgd = "myelo_sp"
dbname ="REACTOME"; gsname = "REACTOME_SIGNALING_BY_NUCLEAR_RECEPTORS%%M27869"  # size 33
dbname ="WIKIPATHWAYS"; gsname = "WP_STRUCTURAL_PATHWAY_OF_INTERLEUKIN_1_IL1%%M39543"  # size 7
tgt = "myelo_m168101";  bgd = "myelo_sp"
```
